---
title: "Paleo superstatistics notebook"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting the data

### PBDB API

To obtain the PBDB data we make use of the API in script `data/pbdb_data_get.R`, which accesses the API and cleans the data by:

- removing poorly lithified specimens
- removing collections at the basin scale
- including only fine-scale stratigraphy (below the "group" level)
- resolving taxonomy to the genus or subgenus level where availible (storing genus or subgenus as `otu`)
- combining multiple records of the same OTU per collection
- importing standardized timebins from \url{fossilworks.org} (timebins are scraped with script `data/fossilworks_tbins_intervals.R`)


```{r code=readLines('data/pbdb_data_get.R'), eval=FALSE}
```


### Scraping fossilworks

The script to pull Alory's time bins (`data/fossilworks_tbins_intervals.R`) is below: 

```{r code=readLines('data/fossilworks_tbins_intervals.R'), eval=FALSE}
```


## Three timer and publication bias correction

Once the data have been downloaded and cleaned, we correct for incomplete and biased sampling with the script `re-do.R`

```{r code=readLines('re-do.R'), eval=FALSE}
```